Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-29 analysis/paper/paper.Rmd
more edits to management implications

We see two potential management actions: 1) removing the Dump Creek delta to restore flow and 2) a local northern pikeminnow bounty program to encourage harvest of northern pikeminnow in Deadwater Slough.

projections for the lower Columbia River (Beamesderfer et al. 1996) where substantial Northern Pikeminnow predation impacts on salmonids have led to bounty programs aimed at reducing Northern Pikeminnow abundance.

<!--
Is there anything in the literature that discusses habitat preferences for pike minnow? If we speed up velocities and add some cover potentially pike minnow predation success will be lowered. Will these fish just move elsewhere? 
-->

>>>>

We envision two potential management actions at reducing predation at Deadwater Slough: 1) removing the Dump Creek alluvial fan and 2) a local Northern Pikeminnow bounty program to encourage harvest in Deadwater Slough aimed at reducing the predator population size. Both actions have evident pros and cons. Removing the Dump Creek alluvial fan has the benefit of restoring the Salmon River to a more natural flow (e.g., increased water velocity, decreased depth) and bathymetry (e.g., heterogeneous substrate) that likely existed in the reach prior to the failure of the mining reservoir dam in Dump Creek. Restoring flow would presumably reduce Northern Pikeminnow densities to levels similar to upstream and downstream reaches where recent studies identified higher survival and transition probabilities than those at Deadwater Slough (@Axel2015; @Ackerman2018; @Porter2019). However, we do not know the feasibility of removing or alleviating the Dump Creek alluvial fan, and if it could be removed or altered, future disturbance events could potentially move additional loose sediment in the Dump Creek drainage, thereby creating a new alluvial fan. Additionally, restoring the flow would likely diminish the value of the reach as an important bird watching and nesting area.

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids elsewhere (citation). A local bounty program in the Salmon River has the potential to boost the local economy (to an unknown degree) which has been negatively impacted by low returns of Chinook Salmon, steelhead, and Sockeye Salmon in recent decades. Many local business rely on license and gear sales for local sport fisheries, but low salmon and steelhead returns have eliminated or reduced the duration of fisheries in some years, and a local Northern Pikeminnow bounty program could allow the local economy to recoup some of those lost revenues. On the downside, bounty programs require continual effort and harvest to suppress the population size and so would require year of continued funding and local buy-in to remain effective. If funding or effort was alleviated, Northern Pikeminnow would likely recover to a higher population size without additional alterations to the habitat. 
<!--
What about adding cover? It could potentially provide refugia for juvenile salmonids, but could make recreational angling more difficult

Is there anything in the literature that discusses habitat preferences for pike minnow? If we speed up velocities and add some cover potentially pike minnow predation success will be lowered. Will these fish just move elsewhere? 
-->
